---
title: "EES 3310/5310 Lab #1"
subtitle: "Exercises from Chapters 2 and 3"
author: "put your name here"
date: "Lab: Monday Aug. 28, 2017. Due: Friday, Sept. 1."
output:
  pdf_document: default
  html_document: default
  github_document: default
---

```{r setup, include=FALSE}
# This section sets up some options on knitr for processing the RMarkdown file
knitr::knit_hooks$set(inline = function(x) { knitr:::format_sci(x, 'md')})
knitr::opts_chunk$set(echo = TRUE)
```
```{r initialize, include=FALSE}
# This section loads necessary R libraries and sources scripts that define 
# useful functions format_md.
# 
library(tidyverse)
library(stringr)

source("scripts/format_md.R")
```
# Instructions:

* Clone a local copy of the repository from github,
* In your local repository, edit `Lab_01_Activities.Rmd` to answer the questions
  in the exercises,
* knit `Lab_01_Activities.Rmd` into a PDF file,
* commit your changes to the local repository, including the edits to the Rmd file
  and the new PDF file,
* and push the changes from your local repository to the master repository at github.

# Exercises from Chapter 2

## Exercise 2.1

Consider exercise 1 in Chapter 2, on p. 17 of our textbook, 
_Global Warming: Understanding the Forecast_:

> A joule (J) is an amount of energy, and a watt (W) is a rate of using energy, 
> defined as 1 W = 1 J/s. How many Joules of energy are required to run a 100-W 
> light bulb for one day?

```{r ex_2_1_a}
seconds_per_hour = 60 * 60 # number of seconds in one hour
hours_per_day = 24 # number of seconds in one day
light_bulb_power = 100
joules_per_day = light_bulb_power * seconds_per_hour * hours_per_day

joules_per_day
```

It can be hard to read a long number like that, so we can use R's format command to add commas:

```{r formatting_example}
format_md(joules_per_day, comma = TRUE)
```

**Answer:** A `r light_bulb_power` watt light bulb burns `r light_bulb_power` Joules per second, 
so the light bulb burns `r light_bulb_power * seconds_per_hour` Joules per hour and
`r joules_per_day` Joules per day.

> Burning coal yields about 30 &times; 10^6^ J of energy per kilogram of coal burned. 
> Assuming that the coal power plant is 30% efficient,
> how much coal has to be burned to light that light bulb for one day?

```{r ex_2_1_b}
joules_heat_per_kg_coal = 30.E+6
efficiency = 0.30
joules_electricity_per_kg_coal = joules_heat_per_kg_coal * efficiency
kg_coal_per_joule_electricity = 1 / joules_electricity_per_kg_coal
kg_coal_per_day = joules_per_day * kg_coal_per_joule_electricity

kg_coal_per_day
```

**Answer:** 1 kilogram of coal produces `r format_md(joules_heat_per_kg_coal, digits = 1)` joules of heat, which is converted into `r format_md(joules_electricity_per_kg_coal, digits = 1)` joules of electricity.
We can turn this around and figure that to get 1 joule of electricity takes 1 / (`r format_md(joules_electricity_per_kg_coal, digits = 1)`) = `r format_md(kg_coal_per_joule_electricity, digits = 1)` kg of coal,
so to get `r format_md(joules_per_day, digits = 1)` joules of electricity to power the light bulb for one day takes `r format_md(kg_coal_per_day, digits = 2)` kg of coal.

## Exercise 2.2

> A gallon of gasoline carries with it about 1.3 &times; 10^8^ J of energy. 
> Given a price of $3 per callon, how many Joules can you get for a dollar?

```{r ex_2_2_a}
# Put your R code here to answer the question
```

**Answer:** You can get ... Joules per dollar from gasoline.

> Electricity goes for about $0.05 per kilowatt hour. A kilowatt hour is just a 
> weird way to write Joules because a watt is a joule per second, and a kilowatt 
> hour is the number of Joules one would get from running 1000 W time one hour
> (3,600 seconds). In the form of electricity, how many Joules can you get for a 
> dollar?

```{r ex_2_2_b}
# Put your R code here to answer the question
```

**Answer:** put your answer here ...

> A standard cubic foot of natural gas carries with it about 1.1 &times; 10^6^ Joules
> of energy. You can get about 5 &times; 10^5^ British Thermal Units (BTUs) of gas
> for a dollar, and there are about 1,030 BTUs in a standard cubic foot. 
> How many Joules of energy in the form of natural gas can you get for a dollar?

```{r ex_2_2_c}
# Put your R code here to answer the question
```

**Answer:** put your answer here ...

> A ton of coal holds about 3.2 &times; 10^10^ J of energy and costs about $40.
> How many Joules of energy in the form of coal can you get for a dollar?

```{r ex_2_2_d}
# Put your R code here to answer the question
```

**Answer:** put your answer here ...

> Corn oil costs about $0.10 per fluid ounce wholesale. A fluid ounce carries
> about 240 dietary Calories (which a scientist would call kilocalories).
> A dietary Calorie is about 4200 J. How many Joules of energy in the form of
> corn oil can you get for a dollar?

```{r ex_2_2_e}
# Put your R code here to answer the question
```

**Answer:** put your answer here ...

> Rank these five energy sources from cheap to expensive. 
> What is the range of prices?

**Answer:** put your answer here ...


## Exercise 2.3 (Grad students only)

> This is one of those job-interview questions to see how creative you are, 
> analogous to one I heard: "How many airplanes are over Chicago at any given time?" 
> You need to make stuff up to get an estimate and demonstrate your management 
> potential. 
> The question is: _What is the efficiency of energy production from growing corn?_
> 
> Assume that sunlight deposits 250 W/m^2^ of energy on a corn field, averaging 
> over the day-night cycle. 
> There are approximately 4,200 J per dietary Calorie. 
> How many Calories of energy are deposited on a square
> meter of field over the growing season? 
>
> (Note: the word "calorie" has two different meanings. Physicists and chemists, 
> use "calorie" (with a lower-case "c") to refer to a thermodynamic unit of heat, 
> but nutritionists use the word Calorie (with a capital 'C')
> to mean 1 kilocalorie (1000 thermodynamic calories), so when you see "Calories" 
> on a food label, it means kilocalories. To keep this exercise simple, I have 
> edited the textbook version so we only need to think in terms of dietary Calories.)


```{r ex_2_3_a}
# Put your R code here to answer the question
```

**Answer:** put your answer here ...

> Now guess how many ears of corn grow per square
> meter, and guess what the number of dietary Calories is that you get for eating 
> an ear of corn. Compare the sunlight energy with the corn energy to get the 
> efficiency.

**Answer:** put your answer here ...

## Exercise 2.4

> The Hoover Dam produces $2 \times 10^{9}$ W of electricity. It is composed of 
> $7 \times 10^{9}$ kg of concrete.
> Concrete requires 1 MJ of energy (1 megajoule, 1,000,000 Joules) to produce 
> per kilogram. 
> How much energy did it take to produce the dam? How long is the "energy payback 
> time" for the dam?

```{r ex_2_4_a}
# Put your R code here to answer the question
```

**Answer:** put your answer here ...

> The area of Lake Mead, formed by Hoover Dam, is 247 mi^2^. Assuming 250 W/m^2^ 
> of sunlight falls on Lake Mead, how much energy could you produce if instead of 
> the lake you installed solar cells that were 12% efficient? 
> (1 mile is 1609 meters; how many square meters are in a square mile?)

```{r ex_2_4_b}
# Put your R code here to answer the question
```

**Answer:** put your answer here ...

## Exercise 2.5

> It takes approximately $2 \times 10^{9}$ J of energy to manufacture 1 m^2^ of 
> crystalline-silicon photovoltaic cell. (Actually, the number quoted was 
> 600 kilowatt hours. Can you figure out how to convert kilowatt hours into 
> Joules?) Assume that the solar cell is 12% efficient, and calculate how long it
> would take, given 250 W/m^2^ of sunlight, for the solar cell to repay the energy 
> it cost for its manufacture.

```{r ex_2_5}
# Put your R code here to answer the question
```

**Answer:** put your answer here ...

## Exercise 2.7

> Infrared light has a wavelength of about 10 $\mu$m. What is its wave number in
> cm^-1^?

```{r ex_2_7_a}
# Put your R code here to answer the question
```

**Answer:** put your answer here ...

> Visible light has a wavelength of about 0.5 $\mu$m. What is its frequency in 
> Hz (cycles per second)?

```{r ex_2_7_b}
# Put your R code here to answer the question
```

**Answer:** put your answer here ...

> FM radio operates at a frequency of about 40 kHz. What is its wavelength?

```{r ex_2_7_c}
# Put your R code here to answer the question
```

**Answer:** put your answer here ...
